\documentclass[10pt]{report}
\setlength{\textheight}{8.0in}
\setlength{\textwidth}{6.00in}
\setlength{\parskip}{3mm}
\setlength{\parindent}{0.0in}
\setlength{\topmargin}{5mm}
%\setlength{\bottommargin}{5mm}
\setlength{\headheight}{5mm}
\setlength{\headsep}{10mm}
\begin{document}
{\large
   M.R.I. SCAN OF LUMBO S{\}ACRAL SPINE \\
   IMAGE PARMETERS \\
   AXIAL 5 mm SPIN ECHO TI, FSE T2 \\
   SATITTAL 5mm SPIN ECHO T1, FSE T2 \\
   FINDINGS \\
   L5-S1 Disc appears Hypointense on T2 weighted image suggesting of Degeneration
   and a right paracentral disc protrusion with caudal migration is seen at this
   level compressing the Issilateral nerve roots. No other obvious disc bulges are
   seen. \\
   Vertebral alignment and marrow signal intensities are maintainer. No focal
   vertebral lesions are seen. No abnormal paraspinal shadows are seen. Spinal
   canal is of normal dimensions. Facet joints and ligamentum plava are
   unremarkable.  \\
   IMPRESSION:MRI findings reveal evidence of degenerative changes in the L5-SI
   disc with a right paracentral disc extrusion and candal migration causing
   ipsilateral nerve root compression. No other significant abnormality is notes. 
   The M.R.I. was consulted from Neuro-Physicions and Neuro-surgeon. The Neuro
   Physician has said that disc is prolapsed. In conventional treatment (i.e.
   without surgery) the chances of cure are about 60%. Neuro-Surgeon has straight
   away advised to undergo Surgery. The Surgery is said take Risky, success rate
   is low. Moreover some side effects like paralytics or other complications are
   also expected. So we have decided to first take treatment of Neuro-Physician.
   He has initially advised complete Bed-Rest for 4 weeks. During Bed-Rest it is
   not even allowed to come down even for the of toilet. \\
}
\end{document}
